---
title: "Tidy Tuesday 2021, Week 9"
description: |
  Using GGplot to make scatterplots looking at gender, race, age and earnings in the US
author:
  - name: Joanna Prout
    url: {}
date: 2021-02-28
output:
  distill::distill_article:
    self_contained: false
---


This week's Tidy Tuesday challenge included data from the Bureau of Labor Statistics (BLS).  The data I selected included a break down of weekly earnings over time from 2010-2020 by gender, race, and age.

Knowing that all of these factors influence how much people are paid, I created two scatterplots (one for men and one for women) of earnings over time with race indicated by point color and age group indicated by shape.


```{r setup, echo = TRUE}
library (tidyverse)
library (lubridate)
library (showtext)
library (RColorBrewer)
font_add_google ("Roboto")
showtext_auto ()
# Get the Data
earn <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-23/earn.csv')
```


```{r, echo = TRUE}
d <- earn %>%
  select (sex, race, age, ethnic_origin, year, quarter, median_weekly_earn) %>%
  filter (age == "16 to 24 years"| age == "25 to 54 years" | age == "55 years and over") %>%
mutate(date = (yq(paste(year, quarter, sep = "-")))) # change year and quarter into date format
```


```{r, echo=TRUE}
plot <- ggplot (d %>% filter (ethnic_origin != "Hispanic or Latino" & race != "All Races" & sex != "Both Sexes"), aes (x=date, y=median_weekly_earn, color = race, shape = age)) +
  geom_point(size = 2) +
  facet_wrap (~sex, ncol = 2)
earnings <- plot + theme (text = element_text (family = "Roboto",  size = 30),
         plot.title = element_text (face = "bold", size = 32),
         legend.position = "bottom",
         legend.key = element_rect (fill = "#e3e4e1"),
         legend.background = element_rect (fill = "#e3e4e1"),
         panel.grid = element_blank (),
         panel.background = element_rect(color = "#e3e4e1", fill = "#e3e4e1"),
         plot.background = element_rect(color = "#e3e4e1", fill = "#e3e4e1"),
         axis.text.y = element_text (size = 20, hjust = 0.5),
         axis.line.x = element_line (),
         axis.line.y = element_line (),
         axis.ticks.y = element_blank (),
         axis.ticks.x = element_blank (),
         legend.box = "vertical",
         plot.margin = margin (0.5, 0.5, 0.5, 0.5, "cm")) +
  labs (title = "Earnings of US Employees by Gender, Race, and Age", 
        y = "Median Weekly Earnings in $", 
        x = " ",
        color = "Race",
        shape = "Age",
        caption = "Data from US BLS | Graphic @joieprout") +
  scale_color_brewer(palette = "Dark2")

ggsave ("earnings.png", height = 6, width = 10, dpi = "retina")
```

```{r}
library (knitr)
include_graphics ("earnings.png")
```
