---
title: "Tidy Tuesday 2021, Week 9"
description:
  "A Tidy Tuesday challenge about race and earnings in the US"
author:
  - name: Joanna Prout
  - date: "`r Sys.Date()`"
output: distill::distill_article
---

This week's Tidy Tuesday challenge included data from the Bureau of Labor Statistics (BLS).  The data I selected included a break down of weekly earnings over time from 2010-2020 by race and gender.

Knowing that both of these factors influence how much people are paid, I created two scatterplots (one for men and one for women) of earnings over time with race indicated by point color.

What struck me looking at these graphs is the strata or layers, which seem to indicate there are groups of higher and lower earners within each race.

```{r setup, echo = TRUE}
library (tidyverse)
library (lubridate)
library (showtext)
library (RColorBrewer)

font_add_google ("Roboto")
showtext_auto ()

# Get the Data
earn <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-23/earn.csv')

```

```{r, echo = TRUE}
d <- earn %>%
  select (sex, race, age, ethnic_origin, year, quarter, median_weekly_earn) %>%
  filter (age == "16 to 24 years"| age == "25 to 54 years" | age == "55 years and over") %>%
mutate(date = (yq(paste(year, quarter, sep = "-")))) # change year and quarter into date format

```

```{r, echo=TRUE}
plot <- ggplot (d %>% filter (ethnic_origin != "Hispanic or Latino" & race != "All Races" & sex != "Both Sexes"), aes (x=date, y=median_weekly_earn, color = race, shape = age)) +
  geom_point(size = 2) +
  facet_wrap (~sex, ncol = 2)

plot + theme (text = element_text (family = "Roboto",  size = 15),
         plot.title = element_text (face = "bold", size = 18),
         legend.position = "bottom",
         legend.key = element_rect (fill = "#e3e4e1"),
         legend.background = element_rect (fill = "#e3e4e1"),
         panel.grid = element_blank (),
         panel.background = element_rect(color = "#e3e4e1", fill = "#e3e4e1"),
         plot.background = element_rect(color = "#e3e4e1", fill = "#e3e4e1"),
         axis.text.y = element_text (size = 10, hjust = 0.5),
         axis.line.x = element_line (),
         axis.line.y = element_line (),
         axis.ticks.y = element_blank (),
         axis.ticks.x = element_blank (),
         legend.box = "vertical",
         plot.margin = margin (0.5, 0.5, 0.5, 0.5, "cm")) +
  labs (title = "Earnings of US Employees by Gender, Race, and Age", 
        y = "Median Weekly Earnings in $", 
        x = " ",
        color = "Race",
        shape = "Age",
        caption = "Data from US BLS | Graphic @joieprout") +
  scale_color_brewer(palette = "Dark2")
```







