---
title: "Tidy Tuesday 2021, Week 7: "
description: |
  Data on Wealth and Income Over Time | Urban Institute & US Census
author:
  - name: Joanna Prout
date: 2021-02-11
output:
  distill::distill_article:
    self_contained: false
---

For this week's Tidy Tuesday challenge we looked at data about wealth and inequality in the United States.  I did a simple area plot of the median wealth of Black, Hispanic, and White families from 1989-2016.  The disparity is huge! The visualization shows that while the wealth of White families increased and decreased as a function of economic changes; the wealth of Blank and Hispanic families remained low.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library (tidyverse)
library (showtext)
library (scales)

## Loading Google fonts (https://fonts.google.com/)
font_add_google("Montserrat")

student_debt <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/student_debt.csv')

income_mean <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_mean.csv')

race_wealth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/race_wealth.csv')
```



```{r med wealth over time by race, fig.height = 6 , fig.width = 8, fig.align = "center", message = FALSE, error = FALSE, echo=TRUE}
median_wealth_race <- race_wealth %>% 
  filter (type == "Median") %>%
  filter (year >= 1989) %>%
  filter (race != "Non-White")

## Automatically use showtext to render text
showtext_auto()

plot <- ggplot (median_wealth_race, 
        aes (x = year, 
             y = wealth_family, 
             group=race, 
             fill = race)) + 
 geom_area(color = "white") +
  labs (title = "Median Wealth of US Families by Race Over Time",
          subtitle = "In Dollars Normalized for 2016",
          caption = "Data from Urban Institute and US Census | Graphic by @joieprout",
          x = " ",
          y = " ") +
  facet_wrap (~race) +
  theme (text = element_text (family = "Montserrat", size = 22),
         plot.title = element_text (face = "bold", size = 24),
         legend.position = "none",
         panel.grid = element_blank (),
         axis.text.x = element_text (size = 14),
         axis.text.y = element_text (size = 14),
         axis.ticks.y = element_blank (),
         axis.ticks.x = element_blank (),
          panel.background = element_rect(color = "gray93", fill = "gray93"),
          plot.background = element_rect(color = "white", fill = "white"),
          strip.background = element_rect(color = "gray80", fill = "gray80"),    
          plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"))+
  scale_y_continuous(label = comma) +
  scale_fill_brewer(palette="Dark2")

plot

ggsave ("wealthplot.pdf", width = 10, height = 6)
```


